Figs for public health journal article April 2017

### 2017-04-21 inst/documents/pub_health_j/pub_health_j_figs.Rmd
figs for pub health jour article

---
title: "Insecticide resistance evolution with mixtures and sequences : a model-based graphical introduction and summary"
output: 
  pdf_document:
    #citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex

#output: word_document
author: "Andy South and Ian Hastings"
date: "`r Sys.Date()`"
bibliography: library.bib
csl: biomed-central.csl
fontsize: 11pt
spacing: double
# below worked in test_linenumbers.Rmd
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
---

```{r, eval=FALSE, include=FALSE}
#!!!! paper protocol
#!!!! might be easier to put text and figs into separate Rmd files after draft stage
#!!!! drafts knit to pdf with pdf_document in YAML and this chunk eval=FALSE
#!!!! submission MS, knit to word_document, YAML, this chunk eval=FALSE, 
#!!!!  in WORD: delete figs, add fig numbers to fig legends, set double space, add line numbers 
#!!!! submission figs, set this chunk to eval=TRUE

# options to create final publication quality figures
# causes the document to fail, but the figs do get stored in pap2figs/
library(knitr)
opts_chunk$set(dev="tiff",
               dev.args=list(compression="lzw"),
               dpi=300,
               cache=FALSE,
               fig.path='pap2figs/')
```

@@

### version 12

###alternative titles:
###Evolution of insecticide resistance with insecticide mixtures and sequences : an accessible model to promote understanding.
###An accessible model to promote understanding of the evolution of insecticide resistance : insecticide effectiveness predicted to be key to the relative performance of mixtures.
###A tutorial, primer ...
###Insecticide effectiveness predicted to be most important factor determining whether insecticide mixtures can slow evolution of resistance.

Target : Malaria journal (other suggestions welcome)

THIS IS A CONFIDENTIAL COPY OF THE MS. Please do not circulate without the permission of one of the authors. Comments and suggestions welcome

## Abstract

_must be 350 words, currently 351_

##### Background
Insecticide resistance threatens the control of vectors and particularly malaria. New insecticides are being developed to address this. Potential strategies for using new insecticides include mixtures, sequences and rotations and there are recommendations for how best to implement these to manage resistance. There is, however, limited recent, accessible modelling work addressing the evolution of resistance under different operational strategies. Previous work concludes that preferred strategies will be situation specific. There is potential to improve the level of mechanistic understanding within the operational community of how insecticide resistance can be expected to evolve in response to different strategies.

This paper develops an accessible, mechanistic picture of how the evolution of insecticide resistance is likely to respond to potential intervention strategies to help guide both management and policy. The aim is to reach an audience unlikely to read a more detailed modelling paper. We use our model to develop a mechanistic understanding of how insecticide resistance is expected to increase with the use of insecticides in isolation, sequence and mixtures. The model flexibly represents two independent genes coding for resistance to two insecticides. We look principally at the ability of the insecticides to kill susceptible mosquitoes, how much resistance counteracts this, the proportion of mosquitoes that are exposed to insecticides and costs of resistance.

##### Results
We use the model to demonstrate the evolution of resistance under different inputs and how this fits with intuitive reasoning about the selection pressure exerted by insecticides. We show that an insecticide used in a mixture, relative to alone, always prompts slower evolution of resistance, but resistance to the two insecticides may evolve more slowly when used in sequence. We show that the ability of insecticides to kill susceptible mosquitoes (effectiveness), has the most influence on whether resistance to two insecticides is likely to arise faster in a mixture or sequence.

##### Conclusions
Our model makes more open the process of insecticide resistance evolution and how it is likely to respond to insecticide use. We provide a simple user-interface allowing further exploration (https://andysouth.shinyapps.io/MixSeqResist1). These tools can contribute to operational decisions in insecticide resistance management.

## Keywords
insecticide resistance; public health; mosquitoes; vector-borne diseases; infectious diseases; malaria; population genetics

## Background

Insecticide resistance is a problem for malaria [@WHO2012][@Ranson2016][@Hemingway2016] other vector borne diseases [@IRAC2011] and agriculture [@FAO2012]. Malaria alone still results in hundreds of thousands of deaths per year. Recent malaria control efforts have centred on treated bed nets and indoor residual spraying, both reliant on insecticides. Treated nets were recently estimated to contribute 68% and indoor residual spraying 13% to averting more than 500 million falciparum malaria cases between 2000 and 2015 [@Bhatt2015]. A recent malaria transmission model [@Churcher2016] predicts that even low pyrethroid resistance is likely to increase malaria incidence in Africa by reducing the performance of bed nets.

The WHO produced a Global Plan for Insecticide Resistance Management in malaria vectors (GPIRM)[@WHO2012] which includes recommendations on operational strategies for managing resistance including the use of insecticide mixtures when they become available.  Efforts are under way to develop new insecticides that will be effective in the light of existing resistance and allow additional options within insecticide resistance management. The Innovative Vector Control Consortium (IVCC) was set up in 2005 to develop new vector control tools and particularly new insecticides to address insecticide resistance in malaria transmitting mosquitoes [@Hemingway2006][@IVCC2016]. Three new insecticides are now in development [@IVCC2016] and likely to be available around 2020 [@Ranson2016]. It is important that decisions about how best to use the new insecticides to delay the onset of resistance are made before insecticides are released [@Hemingway2016].

Modelling studies have investigated the evolution of insecticide resistance in insecticide mixtures including in a public health context e.g. [@Curtis1985][@Mani1985][@Roush1989] but much of the work was done more than 20 years ago and there remained some confusion about the results [@Levick2017]. In an earlier paper [@Levick2017] we described the technical details of a flexible model used to investigate the relative benefits of mixtures and sequences. Here we provide an accessible summary of the model and use selected parameter values to describe mechanistically how the evolution of resistance is influenced by different inputs. This mechanistic understanding can contribute to the debate on the relative merits of different insecticide strategies extending existing frameworks [@Roush1989][@IRAC2011][@FAO2012][@WHO2012].

Our modelling approach focuses on the change in frequency of a single resistance gene for each insecticide. It is likely that genes conferring resistance are present in populations even prior to exposure to novel insecticides thus leading to the potential for selection [@Liu2015].

## Methods

The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One gene or locus is represented for each insecticide. Each individual has two copies of the gene and there are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. Thus individuals can either have both resistance genes (RR homozygous), both susceptible genes (SS) or one of each (SR, heterozygous). The combination of the two genes is termed the genotype. Fitness is the main currency of the model, representing how much each genotype survives and reproduces. The fitness of the SS genotype in absence of the insecticide is set at a maximum value of 1. All other fitnesses are less, determined by alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide, the dominance of the resistance allele and the ability of resistance to restore fitness in the presence of the insecticide (see Table 1).

**Table 1. Parameters inflencing the development of insecticide resistance**

Parameter  | Description
---------------------- | ----------------------------------------------------
1. Effectiveness | proportion of susceptible (SS) insects killed by exposure to insecticide
2. Exposure | proportion of insects exposed to insecticide
3. Resistance restoration  | ability of resistance (RR) to restore fitness of insects exposed to insecticide
4. Dominance of restoration | sets fitness of heterozygous (SR) insects between that of SS & RR in presence of insecticide
5. Frequency  | frequency of resistance alleles within the population
6. Cost of resistance  |  fitness of resistant (RR) insects in absence of insecticide
7. Dominance of cost  | sets fitness of heterozygous (SR) insects between that of SS & RR in absence of insecticide

The calculation of fitness for each genotype when considering just a single insecticide is illustrated in Figure 1. In the model itself this is repeated for both insecticides and the fitnesses multiplied. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS and resistance restoration 'restores' a portion of the fitness for RR. A value of 1 for resistance restoration would bring the fitness of RR back to 1 (equivalent to the unexposed SS), a value of 0 for resistance restoration would lead to the RR having the same fitness as SS in the presence of the insecticide and effectively there would be no resistance. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. The 'selection coefficient' of resistance, that may be seen referred to elsewhere [@Curtis1985], can be calculated by multiplying resistance restoration by insecticide effectiveness. Cost of resistance determines the fitness of RR in the portion of the population not exposed to the insecticide.

Dominance of restoration determines how the fitness for SR sits between that of SS and RR for those that are exposed to the insecticide. For those that are not exposed, fitness of SS is set to 1 by definition and resistance cost determines the fitness of RR. Dominance of cost determines how the fitness for SR sits between that of SS and RR in the absence of the insecticide (Figure 1).

The simulation proceeds through generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness. This acts to make the fitter alleles more common over time. Separate sexes and standard sexual reproduction with recombination is included. There is also the potential to set exposure rates to be different for males and females to represent their different behaviours but we do not explore that here.

So far this description just considers a single insecticide and associated resistance allele. In the model itself a second insecticide and resistance allele are included and overall fitness is calculated by multiplying the two results. This allows the model to represent populations exposed to two insecticides together in a mixture.

The model outputs the change in resistance allele frequencies over time measured in generations. We extract the number of generations at which a resistance allele frequency of 50% is reached and term this 'time-to-resistance'. Using a 50% threshold is consistent with previous modelling analyses e.g. [@Curtis1985][@Birget2015]. Preliminary investigations showed that using other potential thresholds of 25 or 75% gave very similar results.

We investigate three insecticide use strategies.

1. single insecticide
1. two insecticides used in sequence, replacing the first with a second once the frequency of resistance alleles reaches a threshold of 50%
1. two insecticides used in a mixture (concurrently)

The model is implemented in R [@RCoreTeam2016], the code is hosted on Github [@South2017] and we provide an on-line user interface [@South2017a] (Fig 2) enabling the reader to change inputs and run the model themselves.

## Results

### Single insecticide
For single insecticide use, higher values of insecticide effectiveness, exposure, resistance restoration and its dominance all resulted in faster resistance spread (Fig 3 A-D) and thus fewer generations to reach a resistance threshold of 0.5. This makes intuitive sense as all increase the impact of the insecticide and thus increase selection for resistance. Similarly, higher values of starting resistance frequency also resulted in shorter times to resistance thresholds (Fig 4A) through the even simpler mechanism that less change is required to reach the threshold.

Cost of resistance (Fig 4B) and its dominance (not shown) had the opposite effect. Higher values led to a slower development of resistance. Again this is intuitive because costs reduce the advantage of resistance and decrease selection. These effects of inputs on single insecticide use are summarised in Table 2.

### Two insecticides
When two insecticides are used we can compare the relative performance of a mixture with sequential use. Each panel in Figures 5-9 compares mixture and sequence for a single combination of inputs. To help understand the mechanisms influencing this relative performance we start with a single base scenario and demonstrate changing inputs individually. In this base scenario we set effectiveness, exposure, resistance restoration, and dominance at 0.5, the starting frequencies of resistance to 0.01 and cost of resistance to 0. Resistance arises slower for sequential than mixture for this base scenario (Fig 5A). Resistance to both insecticides in the mixture follow the same path and reach the threshold at the same time (as would be expected given that they have identical input parameters). When used in sequence the curve (dashed) for each insecticide individually is steeper than in the mixture but, because they happen one after the other, it takes longer for both to reach the resistance threshold (Fig 5A). Thus the time-to-resistance for the mixture divided by the sequence is less than 1 at 0.8.

#### Insecticide exposure and effectiveness

Increasing exposure (the proportion of insects that come into contact with the insecticides) (Fig 5C) decreases the time-to-resistance for both the sequential and mixture strategy. The effect on the mixture is greater so time-to-resistance remains longer for the sequence (the ratio decreasing from 0.8 to 0.6). In contrast keeping the exposure constant and increasing effectiveness (the proportion of SS insects that are killed by contact) of one of the insecticides (Fig 5B) results in a longer time-to-resistance for the mixture relative to the sequence (and the ratio increasing from 0.8 to 1.2). The longer time-to-resistance for the mixture results from a slower increase in resistance for the less effective insecticide. Resistance to the less effective insecticide in the mixture increases slowly initially and then speeds up after the more effective insecticide has reached the resistance threshold. This points a mechanism whereby the more effective insecticide initially slows the rate of evolution of resistance to the less effective one.

The results is that a more effective insecticide increases time-to-resistance when used in a mixture (compare the solid lines in Fig 5A & B). This is opposite to what happens when used in sequence (compare the red dashed line in Fig 5A & B) and in isolation (Fig 2A), when more effective insecticides shortened times-to-resistance.

In summary increasing effectiveness can favour mixtures moving from column 1 to 2 in Figure 5, and increasing exposure can favour sequences moving from row 1 to 2 in Figure 5. Reduced male exposure with respect to female, as might be expected for male mosquitoes not seeking blood meals, had a similar effect to changing overall exposure. Reducing male exposure (not shown) increased times-to-resistance and favoured mixtures over sequences.

Increasing the effectiveness further of either or both insecticides increases the performance of mixtures over sequences in terms of time-to-resistance. Figure 6A uses the same inputs as Fig 5B. From this scenario a higher effectiveness for insecticide2 (Fig 6C) or insecticide 1 (Fig 6B) or both (Fig 6D) all result in a greater positive difference in time-to-resistance for mixture over sequence. This again points to how more effective insecticides can slow resistance to another when used in a mixture.

#### Resistance restoration and dominance
Increasing resistance restoration or its dominance decreases time-to-resistance for both sequences and mixtures (Fig 7) in the same way that it did for sole use (Fig 3). The result is that increasing either or both of dominance and resistance restoration (Figs 7B,C,D) doesn't change the relative time-to-resistance for mixtures and sequences from that in the base scenario (Fig 7A) where it takes resistance longer to develop for the sequence. So, unlike effectiveness and exposure, the levels of resistance restoration and dominance do not effect whether a mixture or a sequence are likely to be favoured.

#### Starting frequencies of resistance
Changing the starting frequency of resistance had a similar effect on time to resistance for both sequences and mixtures and thus had little effect on their relative performance. For example taking the base scenario and reducing the starting frequency of one resistance allele did not change from sequence being favoured (compare Fig 8C to 8A). Similarly taking a scenario in which time-to-resistance is longer for a mixture (Fig 8B) and decreasing the starting frequency of that resistance allele did not change the fact that the mixture was favoured (Fig 8D).

#### Costs of resistance
Increasing the cost of resistance to one insecticide in a mixture leads resistance to that insecticide to evolve more slowly (compare the solid red line to the solid blue line in Figure 9C, the red line is for the resistance with the higher cost). The resistance with the higher cost also increases more slowly in the mixture relative to when it is used in sequence (compare the solid red line to the dotted red line in Fig 9C). The greater effect of cost on mixtures means the advantage of sequences over mixtures in Fig 9A (and Fig 5A) is removed in Fig 9C. With costs set to 0 time-to-resistance for mixture divided by sequence is 0.8 in Fig 9A and 1 in Fig 9C. At higher insecticide effectiveness the benefit of mixture over sequence is improved as seen by comparing Fig 9D to 9B.

A summary of the effects of inputs on the evolution of resistance for mixtures is given in Table 3 and for the difference between mixtures and sequences in Table 4. The mechanisms identified in the tables are covered further in the discussion.

#### Linkage disequilibrium
It has previously been suggested [@Curtis1985][@WHO2012] that using an insecticide to which there was existing resistance in a mixture with a new insecticide with little resistance could lead to more rapid evolution of resistance to the latter through a process termed linkage disequilibrium. Figure 8 C&D show how our model predicts that linkage disequilibrium does not lead to more rapid evolution of resistance to the newer insecticide. In Figures 8 C&D the newer insecticide with lower starting resistance (in red) leads to faster resistance when used alone (red dotted) than when used in a mixture (red solid).

## Discussion

Resistance responds to insecticide use in the model in ways that we can explain mechanistically through the process of selection. This mechanistic explanation can help us develop a more robust understanding of how we would expect resistance to evolve in the field.

When single insecticide use was investigated (Figs 3-4) higher values of inputs that increased the selective advantage of resistance all led to more rapid evolution resistance, and higher values for cost led to slower evolution of (Fig 4B).

These mechanisms, explaining the response of resistance to single insecticide use under different input values, are summarised in Table 2. Referring to figure 1 can help explain these. Exposure and effectiveness act by decreasing the fitness of all genotypes, dominance of restoration acts by increasing fitness of SR and resistance-restoration acts by increasing fitness of RR & SR. Cost acts by decreasing the fitness of RR and SR for those not exposed.

#### Two insecticides
Using two insecticides either in sequence or in mixtures differs from the use of a single insecticide and, where it differs can again help us to understand the dynamics of resistance evolution. Our base scenario (Fig 5A) shows curves for resistance to both insecticides in the mixture that are shallower than those for sequential use. The shallower curves for the mixture can be explained by each insecticide killing individuals that are resistant to the other insecticide. Thus each insecticide reduces the selection pressure for the increase in resistance to the other and could be said to 'protect' the other. However this protection is not necessarily enough to ensure that resistance arises more slowly in a mixture relative to a sequential strategy. Resistance can evolve more slowly for both insecticides in sequential use, despite being faster for each, because they occur one after the other. In Figure 5A it takes 60 generations for resistance to reach the threshold for each insecticide in the sequence, and 90 when they are applied together in the mixture, so the sequence is still favoured because 2x60 is greater than 90. Incraesing the effectiveness of on insecticide to Fig 5B decreases the added times for the sequence to 90 and increases those for the mixture to 115 thus changing the order and favouring the mixture.

The mechanism for this change can be sen by comparing the shape of the curves in Figs 5A & B. In the mixture, resistance to insecticide 1 (with the increased effectiveness) rises at a similar speed in both figures. However, the increased effectiveness of insecticide 1 causes resistance to insecticide 2 to increase more slowly initially in the mixture. Once insecticide 1 reaches resistance of around 50% at around 50 generations, the curve for insecticide 2 becomes steeper. Once the first insecticide has become ineffective due to resistance it stops killing individuals that are resistant to the second insecticide, loses its protective effect and thus stops slowing the rise in resistance to the second insecticide. A similar effect is visible when the effectiveness of both insecticides are increased (Fig 6C). The identical resistance curves for each insecticide in the mixture are shallower than at the lower effectiveness (Fig 5A) because more individuals resistant to each insecticide are being killed by the other insecticide. In this case the 'protection' given to both insecticides by the other declines at the same rate so we don't see the change in slope seen in Fig 5B.

Fig 5B also demonstrates that when two insecticides have a different effectiveness in a mixture, other parameters being equal, resistance will be expected to increase faster to the more effective insecticide. The more effective insecticide prompts both a) higher selection pressure for it's own resistance and b) greater 'protection' reducing the rise in resistance to the other insecticide.

#### Difference between exposure and effectiveness
It seems initially counter-intuitive that whereas increasing effectiveness slows the development of resistance for mixtures, increasing exposure speeds it up. Increasing the exposure to both insecticides results in a decrease in time-to-resistance (from 90 generations in Fig 5A to 40 generations in Fig 5C) where increasing the effectiveness of both insecticides by the same amount results in a slight increase (from 90 in Fig 5A to 115 in Fig 5B). This contrasts with the identical effects that exposure and effectiveness have on a single insecticide (Figs 3 A,B) and on insecticides in sequence (the red dashed lines are the same in Figs 5B & 5C). This difference, that exposure reduces times-to-resistance for a mixture and effectiveness increases it, was unexpected for us.

#### Costs of resistance
Costs of resistance favour mixtures in these analyses because the evolution of resistance is slowed more by cost in a mixture than in a sequence (compare the red solid lines for a mixture to the red dotted line for a sequence in Fig 9C and D). It seems that the protective effect of the other insecticide in a mixture interacts with the cost of resistance to reduce selection pressure more than in a sequence.

The effects of cost of resistance become more as the frequency of resistance increases. The resistance curves for two insecticides in a mixture that differ only in their cost of resistance, diverge more as the frequency of resistance increases (Fig 9C). This makes intuitive sense, as the frequency of resistance genes increases the impact of fitness costs associated with those genes becomes greater. Thus differences due to cost are less noticeable at low resistance frequencies and become more noticeable at higher resistance frequencies. Thus if a lower resistance threshold was used cost would have less of an effect and if a higher resistance threshold was used cost would have more of an effect. Also note that if the costs are set even higher than we set them here (e.g. above 0.2) they can maintain resistance frequencies below 1 or lead to their decline. Unfortunately, we do not expect such high costs to occur in operational settings.

Relatedly, when there are costs of resistance, the frequency of resistance for insecticides that are not being used is likely to decline. Thus the frequency of resistance for the first insecticide in a sequence, after it has been replaced, may decline to a level where it becomes operationally useful again. As such a repeated sequential strategy could offer benefits over mixtures not considered here. To incorporat this one would need to use a more detailed metric than time-to-resistance.

#### Caveats
Our model examines solely the evolution of resistance and not the control of mosquitoes or malaria. Different insecticide use strategies, as well as effecting the evolution of resistance, will also influence both the numbers of vectors and their propensity to transmit the disease [@Rivero2010]. Insecticide resistance could reduce vector transmission of disease (e.g. by reduced lifespan of the vector) or increase transmission (e.g. by reduced immunity of the vector to the disease) [@Rivero2010]. Our model would predict that the best strategy for limiting the development of resistance in the short term would be to use no insecticide. Of course, using no insecticide could have serious consequences for mosquito and malaria abundance. The challenge is to develop good strategies for delivering mosquito and malaria control in the short term while sustaining their operational capacity in the longer term by restricting the development of resistance. Others are combining models of population genetics and population dynamics to investigate these trade-offs [@Barbosa2017].

In our model we represented insecticide interaction as being additive within the mixture. This is consistent with recent work on aphids [@Taillebois2016], although that work did also indicate some synergistic effects. Our model has the potential to incorporate either synergistic or antagonistic effects too.

The development of insecticide resistance in operational field settings is subject to considerable uncertainty [@Ranson2016]. The model we present here is able to cover the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of parameter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). Thus there is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. This previous analysis also highlighted the relative importance of insecticide effectiveness and exposure in determining the relative performance of mixtures and sequences. In addition we provide an on-line user interface to the model [@South2017a] and enabling readers to investigate model behaviour in areas of parameter space we have not shown.

#### Consistency with previous work 
The predicted responses of resistance presented here can be put into the context of previous work and recommendations. Earlier modelling work has made sometimes contrasting predictions and these can be explained by the details of their structure and input choices. One model predicted that "the use of mixtures is always more effective in delaying the onset of resistance often by many orders of magnitude"[@Mani1985]. That model assumed that SS genotypes were always killed, RR always survived, and that a constant 10% of mosquitoes escaped the insecticide. This would be represented in our model by a very high 'insecticide effectiveness' of 1, a 'resistance restoration' of 1 and an 'exposure' of 0.9. This result is consistent with our prediction that the very high effectiveness is likely to favour mixtures over sequences (e.g. see Fig 6D and Table 4).

A subsequent model [@Roush1998] lead to the following more pessimistic conclusion about the potential of mixtures : “As a result of incomplete coverage and residue decay, the mortality of susceptible homozygotes is rarely consistently high enough for pesticide mixtures to be effective” [@Roush1998]. This "mortality of susceptible homozygotes" is equivalent to our insecticide effectiveness. Again this points to the importance of effectiveness and the question of whether the effectiveness levels needed to make mixtures better are attainable operationally. Our result showing that decreasing exposure favours mixtures over sequences is also consistent with figure 3 from that paper [@Roush1998].

One consistent aspect of the results presented here can seem to contradict published advice and has been questioned when we have presented model results. Our model predicts that lower insecticide effectiveness or exposure will lead to slower spread of resistance in sole use and sequences (Figs 3 & 5, Table 2). This is consistent with expectations that lower effectiveness and exposure would lead to lower mortality of susceptibles and thus lower selection pressure for resistance [e.g. @Roush1998]. This can seem to contradict recommendations that poor implementation of insecticide interventions can promote the development of resistance. e.g. from [@FAO2012] "Certain pest control practices have consistently been shown to exacerbate the loss of susceptible pest populations and the development of resistance. These include: ... - the use of application rates that are below or above those recommended on the label, - poor coverage of the area being treated ...".

Poor implementation would likely cause lower effectiveness of and exposure to insecticides, so how can this increase the potential for resistance when our model suggests it would be decreased ? The explanation for this apparent difference is in the effect of dominance. Poor application can reduce the mortality of heterozygotes (RS) thus effectively increasing the dominance of the resistant allele and increasing its rate of evolution [@Barbosa2011][@FAO2012]. Our model results are consistent with this, showing that increasing dominance of restoration leads to faster spread of resistance (Fig 3C, 7). So, the effect of poor implementation of an insecticide intervention on resistance could be either positive or negative depending on the relative effects on dominance of the resistance gene, insecticide effectiveness and exposure. Similarly, the effects of poor application on the relative benefits of mixtures and sequences will depend on these trade-offs because of our result that effectivenes favours mixtures, exposure favours sequences and higher dominance favours neither (Table 4).

#### Implications

Experience from agriculture where single interventions have been associated with a rapid development of resistance have led for calls for a more Integrated Vector Management composed of a series of partially effective tools [@Thomas2012]. It is suggested that such an integrated approach is more sustainable and 'evolution-proof' [@Thomas2012]. A similar combining of interventions has been advocated to cover mosquitoes exhibiting different behaviours [@Killeen2017]. Mosquitoes with different behaviours e.g. feeding outdoors and/or on animals are likely to result in lower exposures to insecticides in nets or sprayed on walls. This lower exposure could favour mixtures over sequences as we have indicated. In both cases our modelling approach could be used to investigate implications for the development of resistance.

_TODO bits to incorporate into discussion :_

_todo more reference to [@Roush1989] who made many of these points 25 years ago_

_Insecticides with some repellency will reduce exposure to the insecticide and thus, as shown here, are likely to slow the evolution of resistance and favour mictures over sequences. This would need to be balanced with a likely loss in protection to the population as a whole provided by the insecticide [@Birget2015]._

_This model can explore when a mixture strategy may be favoured over a sequence purely for the evolution of resistance. Other factors also need to be considered. On the positive side a mixture is likely to kill more vectors and may also limit their ability to transmit the disease. On the negative side a mixture is more expensive._

_todo, somewhere earlier say good thing about effectiveness & exposure is that they can be modified_

#### Conclusions

We have developed flexible, modern model that can replicate previous modelling work and outputs that help us understand the mechanisms by which different insecticide application strategies may be favoured. For assessing the relative benefits of mixtures and sequences we have demonstrated the likely importance of insecticide effectiveness and exposure and costs of resistance. High insecticide effectiveness favours mixtures through the protective effect of one insecticide on the other. High exposure favours sequences because of the higher selection pressures on both resistances together in a mixture. High costs of resistance can favour mixtures because the selection pressures on both resistances are reduced. Similar processes will occur in other strategies such as rotations and mosaics and we are starting to investigate them. Predictions from the model can be tested against data from trials of new insectides to help advance guidelines on good practice.

### Declarations
TODO see https://malariajournal.biomedcentral.com/submission-guidelines/preparing-your-manuscript/research-article

##### Ethics approval and consent to participate
Not applicable

##### Consent for publication
Not applicable

##### Availability of data and material

##### Competing interests

##### Funding

##### Authors' contributions
AS conceived the paper, re-factored the model code, performed the analyses and wrote the first and subsequent versions of the paper. IH conceived the modelling approach, developed the first version of the model in collaboration with Bethany Levick, explained concepts to AS and reviewed and commented on all manuscript versions.

##### Acknowledgements
Thank you to Bethany Levick who developed the first version of the model code.

**Table 2. Effect of inputs on resistance when insecticides used singly or in sequence**

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | -------- | ---------------------------------------------
1. Effectiveness | faster | reduced fitness of SS in presence of insecticide
2. Exposure | faster | reduced fitness of SS overall
3. Dominance of restoration | faster | increased fitness of SR in presence of insecticide
4. Resistance restoration | faster | increased fitness of RR in presence of insecticide
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of RR in absence of insecticide

**Table 3. Effect of inputs on resistance when insecticides used in a mixture**

Parameter to increase  | effect on resistance evolution | Mechanism
----------------------- | ---------- | ---------------------------------------------
1. Effectiveness | **slower** | one insecticide reduces the fitness of individuals resistant to the other thereby reducing selection pressure from the other
2. Exposure | faster (but less than for single) | reduced fitness of individuals susceptible to one insecticide increases selection pressure for that resistance. However at the same time selection pressure is reduced by reduced fitness of resistant individuals caused by the other insecticide
3. Dominance of restoration | faster | increased fitness of heterozygotes
4. Resistance restoration | faster | increased fitness of resistants
5. Frequency  | faster | less change needed to reach resistance threshold
6. Cost of resistance | slower | reduced fitness of resistants in absence of insecticide

**Table 4. Effect of inputs on the difference between mixture and sequential use**

Parameter to increase  | increase favours mix or sequence | Mechanism
----------------------- | -------- | ---------------------------------------------
1. Effectiveness | mixture | Higher effectiveness gives faster resistance for sequence and slower resistance in mixture
2. Exposure | sequence | Higher exposure gives faster resistance for sequence and mixture but the greater effect on mixture favours sequence.
3. Dominance of restoration | neither | Higher dominance gives faster resistance in both sequences and mixtures such that the difference between them is not changed. 
4. Resistance restoration | neither | As for dominance of restoration. Higher resistance restoration gives faster resistance in both sequences and mixtures such that the difference between them is not changed.
5. Frequency  | neither | As for dominance and resistance restoration. Higher starting frequencies give faster resistance in both sequences and mixtures such that the difference between them is not changed.
6. Cost of resistance | mixture | Higher costs slow the evolution of resistance more in a mixture than when used in sequence (but with higher costs in a sequence there is a greater chance for resistance levels to decline for the insecticide not being used).

## Figures

@@

plot_fit_calc( effectiveness=0.8, resistance_restoration=0.5, dominance_restoration=0.6, dominance_cost=0.6, cost=0.3 )

@@

```{r Fig3, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7.5, fig.width=7.5, fig.show='hold', fig.align='left', fig.cap="Single insecticide use and the effect of inputs on changing resistance frequency over time. A. Effectiveness, B. Exposure, C. Dominance of restoration, D. Resistance restoration. Increasing any of the inputs (going from red to blue) results in shorter times-to-resistance. For each panel the chosen input was varied in isolation with the remaining inputs set to 0.5, except for starting frequency of resistance which was set to 0.01). "}

  library(cowplot)

  #list to put plots in for cowplot
  plotlist <- list(4)

  #A effectiveness  
  a <- setExposure(exposure=0.5, insecticideUsed='insecticide1')

  i1 <- setInputOneScenario( max_gen = 500,
                      h.RS1_A0 = 0.5, 
                      h.RS2_0B = 0.5,
                      a = a,
                      phi.SS1_A0 = 0.2,
                      phi.SS2_0B = 0.5,
                      rr_restoration_ins1 = 0.5,
                      rr_restoration_ins2 = 0.5 )

  i2 <- setInputOneScenario( max_gen = 500,
                      h.RS1_A0 = 0.5, 
                      h.RS2_0B = 0.5,
                      a = a,
                      phi.SS1_A0 = 0.5,
                      phi.SS2_0B = 0.5,
                      rr_restoration_ins1 = 0.5,
                      rr_restoration_ins2 = 0.5 )

  i3 <- setInputOneScenario( max_gen = 500,
                      h.RS1_A0 = 0.5, 
                      h.RS2_0B = 0.5,
                      a = a,
                      phi.SS1_A0 = 0.8,
                      phi.SS2_0B = 0.5,
                      rr_restoration_ins1 = 0.5,
                      rr_restoration_ins2 = 0.5 )

  #input <- cbind(input, inputOneScenario)
  input <- cbind(i1,i2,i3)

  listOut <- runModel2( input )

  df_resist <- get_resistance(locus=1, listOut)

  #A  
  plotlist[[1]] <- ggplot(df_resist, aes(x=generation, y=resistance, colour=factor(phi.SS1_A0))) + 
         theme_bw() +  
         theme(legend.position = "bottom", legend.key = element_blank()) + 
         guides(colour = guide_legend(reverse=TRUE)) +  
         labs(colour = "effectiveness") +
         coord_trans(y = "log10") +    
         geom_line()

  #B exposure
  exposure <- 0.2
  a <- setExposure(exposure = exposure, insecticideUsed='insecticide1')

  i1 <- setInputOneScenario( max_gen = 500,
                      h.RS1_A0 = 0.5, 
                      h.RS2_0B = 0.5,
                      a = a,
                      exposure = exposure, #used for post-run analysis
                      phi.SS1_A0 = 0.5,
                      phi.SS2_0B = 0.5,
                      rr_restoration_ins1 = 0.5,
                      rr_restoration_ins2 = 0.5 )

  exposure <- 0.5
  a <- setExposure(exposure = exposure, insecticideUsed='insecticide1')

  i2 <- setInputOneScenario( max_gen = 500,
                      h.RS1_A0 = 0.5, 
                      h.RS2_0B = 0.5,
                      a = a,
                      exposure = exposure, #used for post-run analysis
                      phi.SS1_A0 = 0.5,
                      phi.SS2_0B = 0.5,
                      rr_restoration_ins1 = 0.5,
                      rr_restoration_ins2 = 0.5 )

  exposure <- 0.8
  a <- setExposure(exposure = exposure, insecticideUsed='insecticide1')

  i3 <- setInputOneScenario( max_gen = 500,
                      h.RS1_A0 = 0.5, 
                      h.RS2_0B = 0.5,
                      a = a,
                      exposure = exposure, #used for post-run analysis
                      phi.SS1_A0 = 0.5,
                      phi.SS2_0B = 0.5,
                      rr_restoration_ins1 = 0.5,
                      rr_restoration_ins2 = 0.5 )

  input <- cbind(i1,i2,i3)

  listOut <- runModel2( input )

  df_resist <- get_resistance(locus=1, listOut)

  #B  
  plotlist[[2]] <- ggplot(df_resist, aes(x=generation, y=resistance, colour=factor(exposure))) + 
         theme_bw() +
         theme(legend.position = "bottom", legend.key = element_blank()) + 
         guides(colour = guide_legend(reverse=TRUE)) +
         labs(colour = "exposure", y=NULL) +
         coord_trans(y = "log10") +   
         geom_line()

  #plot_grid( plotlist[[1]],plotlist[[2]],ncol=2, rel_widths=c(1,1), labels='AUTO')

  #C single insecticide dominance
  a <- setExposure(exposure=0.5, insecticideUsed='insecticide1')

>>>>

---
title: "Figs for public health journal article April 2017"
output: 
  pdf_document:

#output: word_document
author: "Andy South and Ian Hastings"
date: "`r Sys.Date()`"
---

```{r, eval=FALSE, include=FALSE}
# options to create final publication quality figures
# causes the document to fail, but the figs do get stored in figs/
library(knitr)
opts_chunk$set(dev="tiff",
               dev.args=list(compression="lzw"),
               dpi=300,
               cache=FALSE,
               fig.path='figs/')
```

@@

@@

plot_fit_calc( simple=TRUE, effectiveness=0.8, resistance_restoration=0.5, dominance_restoration=0.6, dominance_cost=0.6, cost=0.3 )

@@

#TODO decide which of later figures to modify for pubhealth J paper
